---
title: "R Coding Exercise"
---

Placeholder file for the future R coding exercise.

```{r}
#load dslabs package
library("dslabs")
#look at help file for gapminder data
help(gapminder)
library(tidyverse)
library(dplyr)
library(ggplot2)

```

starting httpd help server .... done

```{r}
#get an overview of data structure
str(gapminder)

```
```{r}
#get a summary of data
summary(gapminder)

```

```{r}
#determine the type of object gapminder is
class(gapminder)
```


```{r}
#Assigning only the African countries to a new object/variable called africadata
africadata <- gapminder[gapminder$continent == "Africa", ]
```


```{r}
#Running str for africadata
str(africadata)
```

```{r}
#Summary of africadata
summary(africadata)
```
```{r}
#Creating africa_infant_life which is a subset of africadata containing only 'infant_mortality' and 'life_expectancy'

africa_infant_life <- africadata[,c("infant_mortality", "life_expectancy")]
```


```{r}
#Running str for africa_infant_life
str(africa_infant_life)
```
```{r}
#Summary of africa_infant_life
summary(africa_infant_life)
```
```{r}
#Creating africa_pop_life which is a subset of africadata containing only population and life_expectancy

africa_pop_life <- africadata[,c("population", "life_expectancy")]

```

```{r}
#Running str for africa_pop_life
str(africa_pop_life)
```

```{r}
#Summary of africa_pop_life
summary(africa_pop_life)
```


Plotting life expectancy as a function of infant mortality.

```{r}
p1 <- africa_infant_life %>% ggplot(aes(x=infant_mortality, y=life_expectancy)) + geom_point() + geom_smooth(method='lm')
plot(p1)

```
The figure exhibits a negative correlation between infant mortality and life expectancy. 


```{r}
#Plotting life expectancy as a function of log of population for the africa_pop_life
p2 <- africa_pop_life %>% ggplot(aes(x=log(population), y=life_expectancy)) + geom_point() + geom_smooth(method='lm')
plot(p2)

```

The above figure indicates a positive correlation between log of population and life expectancy.The figure has streaks of data which appears to go together. It is because africadata contains multiple years of data on population and life expectancy for multiple countries. Each 'streak' could represent  yearly growth in life expectancy and population of individual countries.      
```{r}
#Exploring which years have missing data for infant mortality

missing_infmort_years <- africadata %>%
  filter(is.na(infant_mortality)) %>%
  select(year) %>%
  distinct() %>%
  arrange(year)

print(missing_infmort_years)


```
Perform some analysis on the Year 2000 data 

```{r}
#Creating africadata_2000 which is a subset of africadata containing data for the year 2000 only. 

africadata_2000 <- africadata %>% filter(year == 2000)
```


```{r}
#Running str for africadata_2000
str(africadata_2000)
```
```{r}
#Running summary for africadata_2000
summary(africadata_2000)
```


```{r}
#Plotting life expectancy as a function of infant mortality for the africadata_2000.

p3 <- africadata_2000 %>% ggplot(aes(x=infant_mortality, y=life_expectancy)) + geom_point() + geom_smooth(method='lm')
plot(p3)

```
The above plot exhibits a negative correlation between infant mortality and life expectancy in  the African countries during the year 2000.


```{r}
#Plotting life expectancy as a function of log of population for the africadata_2000

p4 <- africadata_2000 %>% ggplot(aes(x=log(population), y=life_expectancy)) + geom_point() + geom_smooth(method='lm')
plot(p2)

```
There is still visible correlation between population size and life expectancy of the African countries in the year 2000.

Now, we fit some simple models on the Year 2000 data (africadata_2000).
First we fit linear model (fit1) using life expectancy as outcome and infant mortality as predictor
```{r}
fit1 <- lm(life_expectancy ~ infant_mortality, africadata_2000)  

```

```{r}
#Printing the results of the linear model fit1
summary(fit1)
```
Based on the p-value of fit1 which is less than 0.001, infant_mortality is a significant predictor of life expectancy in the African countries. One infant mortality decreases the life expectancy by approximately 0.19 years on an average.   


Next, we fit linear model (fit2) using life expectancy as outcome and population as predictor
```{r}
fit2 <- lm(life_expectancy ~ population, africadata_2000)  

```

```{r}
#Printing the results of the linear model fit2
summary(fit2)
```

The p-value (0.616) of population in the model fit2 is larger than 0.1. Hence, population is not a significant predictor of life expectancy. 








































































